**OptiClust: Improved method for assigning amplicon-based sequence data to operational taxonomic units**

### 2016-10-25 submission/Westcott_OptiClust_mSystems_2016.Rmd
First full draft of Introduction

Amplicon-based sequencing has provided incredible insights into Earth's microbial biodiversity.

Numerous methods have been proposed

Needed an objective criteria to evaluate the various methods methods

Average neighbor is consistently the best method, but others including USEARCH and VSEARCH resulted in comparable results

Difficulty with amount of memory and time required to complete average neighbor algorithm is a significant hurdle to completion of the method.

Instead of retrospectively evaluating methods for their ability to correctly assign sequences we sought to develop a method that would prospectively assign sequences to OTUs to optimize classification metrics.

Clustering quality within the algorithm is assessed by counting the number of true positives (TP), true negatives (TN), false positives (FP), and false negatives (FN) based on the pairwise distances. Sequence pairs that are within the user-specified threshold and are clustered together represent TPs and those in different OTUs are FNs. Those sequence pairs that have a distance larger than the threshold and are not clustered in the same OTU are TNs and those in the same OTU are FPs. These counts are used to calculate the optimization metric.

***Optimization of clustering using composite metrics significantly improves clustering quality (Figure 2, 3, 4).*** There are multiple metrics available to assess clustering quality. The mothur-based implementation of OptiClust allows the user to use the MCC, F1 score, accuracy, sensitivity, specificity, and the sum of true positives and negatives. The MCC, F1 score, and accuracy are preferred because each of them incorporates all four values from the confusion matrix while others only utilize two values. It is relatively straightforward to implement other metrics. **Comparison of MCC, accuracy, and F1 score to each other for each dataset using the full set of sequences.** Each metric outperformed the observed values for the other clustering algorithms indicating that regardless of the metric one uses to evaluate clustering quality, the OptiClust algorithm generates better OTU assignments than any of the other methods. The number of OTUs generated by an algorithm is often used as a metric for clustering quality. We did not observe a significant correlation between clustering quality as measured by MCC, F1 score, or accuracy and the number of OTUs generated by each algorithm. **Interestingly, the values of the MCC, F1 score, and accuracy generated when using each metric to optimize clustering were similar across implementations.** Based on these results and its previous use in the OTU assignment literature, the remainder of our analysis uses the MCC metric for optimization.

@@

The cluster optimization procedure is dependent on the metric that is chosen for optimization. We employed the MCC because it includes the four values from a confusion matrix. Other algorithms such as the furthest neighbor and nearest neighbor algorithms minimize the number of FP and FN, respectively; however, these suffer because the number of FN and FP are not controlled. Alternatively, one could optimize based on the sensitivity, specificity, or accuracy, which are each based on two values from the confusion matrix or they could optimize based on the F1 score, which is based on three values from the confusion matrix. Because these metrics do not balance all four parameters equally, it is likely that one parameter will dominate in the optimization procedure. For example, optimizing for sensitivity could lead to a large number of FPs. Since we would like to minimize both FPs and FNs and not just the total number of false assignments, we decided to optimize utilizing the MCC. The MCC measures the correlation between observed and predicted classifications and is robust to cases where there is an uneven distribution across the confusion matrix. It is possible that other metrics could be developed and employed for optimization of the clustering.

>>>>

Amplicon-based sequencing has provided incredible insights into Earth's microbial biodiversity. It has become common for studies to include sequencing millions of 16S rRNA gene sequences across hundreds of samples. This is three to four orders of magnitude greater sequencing depth than was previously achieved using Sanger sequencing. The increased sequencing depth has highlighted the limitations of existing bioinformatics tools for analyzing these sequences. Furthermore, these sequencing efforts have revealed novel taxonomic diversity that are not adequately represented in reference databases. Because of this, a critical component to the analysis of microbial biodiversity has been the assignment of 16S rRNA gene sequences and other marker gene sequences to operational taxonomic units (OTUs) without reliance on taxonomic classification.

A growing number of algorithms have been developed to assign sequences into OTUs and these can be clustered into three general categories. The first category of algorithms are classification-based and have been termed closed-reference or phylotyping. Sequences are compared to a reference collection and clustered based on the reference sequence that they are similar to. This approach is fast; however, struggles when a sequence is similar to multiple reference sequences that may have different taxonomies. The second category of algorithm have been called *de novo* because they assign sequences to OTUs without the use of a reference. These include hierarchical algorithms such as nearest, furthest, and average neighbor and methods that employ heuristics such as abundance or distance-based greedy clustering as implemented in USEARCH or VSEARCH, Sumaclust, OTUCLUST, and Swarm. These methods tend to be more computationally intense and it has proven difficult to know which method generates the best assignments. A third category of algorithm is a hybrid approach or open-reference clustering. Here sequences are assigned to OTUs using closed-reference clustering and sequences that are not within a threshold of a reference sequence are then clustered using a *de novo* approach. This category blends the strengths and weaknesses of the other method and adds the complication that closed-reference and *de novo* clustering use different OTU definitions.

Several metrics have emerged for assessing the quality of OTU assignment algorithms. These have included the time and memory required to run the algorithm, agreement between OTU assignments and the sequences' taxonomy, sensitivity of an algorithm to stochastic processes, and the number of OTUs generated by the algorithm. Unfortunately, these methods fail to directly quantify the quality of the OTU assignments. An algorithm may complete with minimal time and memory requirements or generate an idealized number of OTUs, but the composition of the OTUs could be incorrect. These metrics also tend to be subjective. For instance, a method may appear to be recapitulate the taxonomy of a synthetic community with known taxonomic structure, but do a poor job when applied to real communities with poorly defined taxonomic structure or for sequences that are prone to misclassification. As an alternative, we developed an approach to objectively benchmark the clustering quality of OTU assignments. This approach counts the number of true positives (TP), true negatives (TN), false positives (FP), and false negatives (FN) based on the pairwise distances. Sequence pairs that are within the user-specified threshold and are clustered together represent TPs and those in different OTUs are FNs. Those sequence pairs that have a distance larger than the threshold and are not clustered in the same OTU are TNs and those in the same OTU are FPs. These values can be synthesized into a single correlation coefficient, the Matthews' Correlation Coefficient (MCC), which measures the correlation between observed and predicted classifications and is robust to cases where there is an uneven distribution across the confusion matrix.

We previously used this approach to assess the OTU assignments of a variety of algorithms using synthetic and real datasets. Consistently, the average neighbor algorithm was identified as among the best or the best algorithm. The distance-based greedy clustering as implemented in VSEARCH has also performed well. The computational resources required to complete the average neighbor algorithm can be significant for large datasets and so there is a need for an algorithm that efficiently produces consistently high quality OTU assignments. Previously, we assessed the quality of the clusters after the completion of the algorithm or post hoc. In the current study we developed and benchmarked a new de novo clustering algorithm. Instead of assessing the methods post hoc, the new method uses a priori calculation of the MCC to direct the algorithm. The result is the OptiClust algorithm, which produces significantly better sequence assignments while making efficient use of computational resources.

@@

The cluster optimization procedure is dependent on the metric that is chosen for optimization. We employed the MCC because it includes the four values from a confusion matrix. Other algorithms such as the furthest neighbor and nearest neighbor algorithms minimize the number of FP and FN, respectively; however, these suffer because the number of FN and FP are not controlled. Alternatively, one could optimize based on the sensitivity, specificity, or accuracy, which are each based on two values from the confusion matrix or they could optimize based on the F1 score, which is based on three values from the confusion matrix. Because these metrics do not balance all four parameters equally, it is likely that one parameter will dominate in the optimization procedure. For example, optimizing for sensitivity could lead to a large number of FPs. Since we would like to minimize both FPs and FNs and not just the total number of false assignments, we decided to optimize utilizing the MCC. It is possible that other metrics could be developed and employed for optimization of the clustering.
